\dppTitle

### 2014-05-31 manuscripts/dpp-msbayes/dpp-msbayes-main.tex
More editing of main text and abstract of dpp ms.

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\cite{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and lineage coalescence.
\change{
More specifically, the sample space of such a model-choice procedure could
include all models ranging from a single divergence-time parameter (i.e.,
simultaneous divergence of all co-distributed taxa) to the fully generalized
model in which each taxon diverged at a unique time.}

@@

\change{
\msb has been used to address biogeographical questions in variety of empirical
systems. Some examples include
(1) whether rising of the Isthmus of Panama caused co-divergence across species
of echinoids co-distributed across the Pacific and Atlantic sides of the ishmus
\cite{Hickerson2006}, 
(2) if an historical seaway across the Baja Peninsula caused co-divergence
across species of squamates and mammals co-distributed both north and south of
the putative seaway \cite{Leache2007},
(3) if species of gall-wasps and their associated parasitoids share divergences
across putative glacial refugia \cite{Stone2012}, and
(4) whether repeated coalescence and fragmentation of the oceanic Islands of
the Philippines during by Pleistocene sea-level fluctuations caused
diversification of vertebrate taxa distributed across the islands
\cite{Oaks2012}.
Such applications of the method often result in strong posterior support for
co-divergence among some of the taxa investigated (e.g.,
\cite{Barber2010,Carnaval2009,Chan2011,Hickerson2006,Leache2007,Plouviez2009,Stone2012,Voje2009,Oaks2012}).
}

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
\change{
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, which can result in small marginal likelihoods relative to models with
fewer divergence-time parameters \cite{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \cite{Oaks2012,Oaks2014reply}.
This serves as a ``natural'' penalty for over-parameterization, which is a
strength of Bayesian model choice.
However, given the sensitivity of marginal likelihoods to the prior, care is
needed when selecting prior distributions \cite{Lindley1957}.
Selecting distributions that are broad or very differently shaped than the
likelihood density (e.g., a uniform prior for divergence times) can lead to
small marginal likelihoods of parameter-rich models even if they are correct.
}

@@

In this study, I introduce a new method, implemented in the software
\dppmsbayes, that extends the model of \msb.
I use this method to test whether alternative parameterizations and priors
improve the behavior of the approximate-Bayesian model-choice approach to
estimating shared divergence events.
The new approach uses a Dirichlet-process prior (DPP) over all possible models
of divergence, and gamma and beta probability distributions in place of uniform
priors on many of the model's parameters.
Using simulations, I show that the new implementation has improved robustness,
accuracy, and power compared to the original model.
Also, my results confirm that the improved performance of the new model
is due to a combination of
(1) more flexible priors on divergence-time and demographic parameters that
avoid placing high prior density in improbable regions of parameter space, and
(2) a diffuse Dirichlet-process prior that does not strongly disfavor
divergence models with intermediate numbers of divergence events.
% Also, our results confirm that both the uniform priors on demographic and
% divergence-time parameters and the U-shaped prior on divergence models
% contribute to the tendency of \msb.
After reanalyzing sequence data from 22 pairs of taxa from the Philippines
\cite{Oaks2012} under the new model, I find a large amount of posterior
uncertainty in the number of divergence events shared among the taxa; a result
in contrast with the original \msb model and congruent with
intuition given the richness of the model and the relatively small amount of
information in the data.

@@

\subsection{The model}
In this section, I describe the model, which is a modification of the model
implemented in \msb \cite{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{\url{https://github.com/joaks1/dpp-msbayes}}).
To perform the analyses described below, I used the freely avaliable,
open-source software package \pymsbayes
(\href{https://github.com/joaks1/PyMsBayes}{\url{https://github.com/joaks1/PyMsBayes}}),
which is a Python API that provides a multi-processing interface to \msb and
\dppmsbayes.
I performed the work described below following the principles of
Open Notebook Science.
Using version-control software, I make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.
I follow much of the notation of Oaks et al.\ \cite{Oaks2012}, but modify it to
aid in the description of the new model.
A summary of my notation can be found in Table \ref{tabNotation}.

@@

\begin{linenomath}
To avoid calculating the likelihood terms of Equation \ref{eq:fullModelCompact},
I distill each sequence alignment \alignment{}{} into a vector of insufficient
summary statistics \alignmentSS{}{}, thus replacing the full dataset
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
with vectors of summary statistics for each alignment
$\ssVectorObs = (\alignmentSSObs{1}{1}, \ldots,
    \alignmentSSObs{\npairs{}}{\nloci{\npairs{}}})$.
\change{
Optionally, for each population pair, the means of the summary statistics can
be calculated across the \nloci{} loci, and the vector can be further reduced
to $\ssVectorObs = (\alignmentSSObs{1}{}, \ldots,
\alignmentSSObs{\npairs{}}{})$.
Either way, with \ssVectorObs in hand, we can estimate the approximate joint
posterior distribution}
{\small
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \ssSpace, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\ssSpace \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\ssSpace \given \hkyModelVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)},
    \label{eq:approxModelCompact}
\end{equation}
}
where \ssSpace is the multidimensional Euclidean space around the vector of
summary statistics, the radius of which is the tolerance \tol.
The sources of approximation are the insufficiency of the statistics and the
\tol being greater than zero.
I describe the full model in detail before delving into the numerical
method of estimating the approximate model.
\end{linenomath}

\subsubsection{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The likelihood terms of Equation \ref{eq:fullModelCompact} can
be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$, is the
probability of the sequence alignment of locus $j$ for population pair $i$
given the gene tree and \hky \cite{HKY} substitution model parameters
\cite[i.e., the ``Felsenstein likelihood'']{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters: The
mutation-rate-scaled effective sizes ($\myTheta{} = 4N\mutationRate$) of the
ancestral, \ancestralTheta{}, and descendant populations, \descendantTheta{1}{}
and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \cite{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

\subsubsection{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate multipliers
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be non-zero, the
prior term $p(\recombinationRate)$ would be added to Equation
\ref{eq:fullModelCompact}, and the prior would be $\recombinationRate \sim
Gamma(\gammaShape{\recombinationRate}, \gammaScale{\recombinationRate})$,
\change{where \gammaShape{\recombinationRate} and
\gammaScale{\recombinationRate} are specified by the investigator.}
\end{linenomath}

\begin{linenomath}
The prior term for the demographic parameters, $p(\demographicParamVector)$,
expands out into its components and as a product over the \npairs{}
pairs of populations
\begin{equation}
    p(\demographicParamVector) =
    \prod_{i=1}^{\npairs{}}
    p(\ancestralTheta{i})
    p(\descendantTheta{1}{i})
    p(\descendantTheta{2}{i})
    p(\bottleScalar{1}{i})
    p(\bottleScalar{2}{i})
    p(\bottleTime{i})
    p(\migrationRate{i}).
    \label{eq:demographicPrior}
\end{equation}
The priors for the demographic parameters are:
$\ancestralTheta{} \sim Gamma(\gammaShape{\ancestralTheta{}},
\gammaScale{\ancestralTheta{}})$,
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\bottleScalar{1}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleScalar{2}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleTime{} \sim U(0, 1)$,
and
$\migrationRate{} \sim Gamma(\gammaShape{\migrationRate{}},
\gammaScale{\migrationRate{}})$.
\change{
By default, \ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}
share the same prior (i.e., $\gammaShape{\ancestralTheta{}} =
\gammaShape{\descendantTheta{}{}}$ and $\gammaScale{\ancestralTheta{}} =
\gammaScale{\descendantTheta{}{}}$), but a separate gamma-distributed prior can
be assigned to \ancestralTheta{}.
Also, the \bottleScalar{1}{}, \bottleScalar{2}{}, and \migrationRate{}
parameters are optional (i.e., it can be assumed that there is no migration
between populations of each pair, and that the population size of each
descendant population has been constant through time).
For the parameters included in the model, the hyper-parameters of the prior
distributions are specified by the investigator.}
\end{linenomath}

@@

\subsubsection{Time scale}
As described above, divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by Oaks et al.\ \cite{Oaks2012}, in \msb, \myTheta{C} is half of
the upper limit of the continuous uniform prior on the mean of the descendant
population sizes.
This is only equal to the expectation of the prior if the lower limit is zero.

\subsection{ABC estimation of the posterior of the model}
\subsubsection{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
I use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
\change{
(1) drawing values of all the model's parameters, which I denote \hpvector{},
from their respective prior distributions;
% (2) rescaling the divergence times
% $\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
% from units proportional to time to units proportional to the expected number
% of mutations via Equation \ref{eq:divTimeScaled} to get 
% $\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
% \divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(2) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(3) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
and
(4) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments;
Optionally, an additional step can be performed to reduce the summary
statistics to the means across loci for each population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$.
Either way, \ssVector{} contains the same summary statistics as those estimated
from the observed data \ssVectorObs.
}
After repeating this procedure \numPriorSamples times, I obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

>>>>

This inference problem is challenging due to the stochastic nature by which
mutations occur in populations and how they are inherited over generations
\cite{Hudson1990,WakeleyCoalescent}.
Thus, a method for estimating historical patterns of divergences across taxa
should explicitly model the stochastic mutational and ancestral processes that
generate and filter the genetic variation we observe in present-day genetic
data.
An appealing approach would be a comparative, Bayesian model-choice method for
inferring the probability of competing divergence histories while integrating
over uncertainty in mutational and ancestral processes via models of nucleotide
substitution and lineage coalescence.
\change{
The sample space of such a model-choice procedure would include all models
ranging from a single divergence-time parameter (i.e., simultaneous divergence
of all co-distributed taxa) to the fully generalized model in which each taxon
diverged at a unique time.}

@@

\change{
\msb has been used to address biogeographical questions in variety of empirical
systems. Some examples include
(1) whether the rise of the Isthmus of Panama caused co-divergence among
species of echinoids co-distributed across the Pacific and Atlantic sides of
the isthmus \cite{Hickerson2006}, 
(2) if an historical seaway across the Baja Peninsula caused co-divergence
across species of squamates and mammals co-distributed both north and south of
the putative seaway \cite{Leache2007},
(3) if species of gall-wasps and their associated parasitoids share divergences
across putative glacial refugia \cite{Stone2012}, and
(4) whether repeated coalescence and fragmentation of the oceanic Islands of
the Philippines during Pleistocene sea-level fluctuations caused
diversification of vertebrate taxa distributed across the islands
\cite{Oaks2012}.
Such applications of the method often result in strong posterior support for
co-divergence among some of the taxa investigated (e.g.,
\cite{Barber2010,Carnaval2009,Chan2011,Hickerson2006,Leache2007,Plouviez2009,Stone2012,Voje2009,Oaks2012}).
}

For priors on divergence-time and demographic parameters, \msb uses continuous
uniform probability distributions.
\change{
This causes divergence models with more divergence-time parameters to integrate
over a \emph{much} greater parameter space with low likelihood yet high prior
density, which can result in small marginal likelihoods relative to models with
fewer divergence-time parameters \cite{Jeffreys1939,Lindley1957}.
Given that the marginal likelihood of a model weighted by its prior is what
determines its posterior probability, this can cause support for models with
fewer divergence events \cite{Oaks2012,Oaks2014reply}.
% The use of marginal likelihoods to estimat model probabilities provides a
% ``natural'' penalty for overparameterization, and can be a great strength of
% the Bayesian approach.
This is not a critique of Bayesian model choice in general; comparing models by
their marginal likelihoods provides a ``natural'' penalty for
overparameterization and can be a great strength of the Bayesian approach.
However, given the sensitivity of marginal likelihoods to the prior, care is
needed when selecting prior distributions \cite{Lindley1957}.
Selecting distributions that will often place high prior density in large
regions of parameter space with low likelihood can lead to small marginal
likelihoods of parameter-rich models even if they are correct.
}

@@

In this study, I introduce a new method, implemented in the software
\dppmsbayes, that extends the model of \msb.
I use this method to test whether alternative parameterizations and priors
improve the behavior of the approximate-Bayesian model-choice approach to
estimating shared divergence events.
The new approach uses a Dirichlet-process prior (DPP) over all possible models
of divergence, and gamma and beta probability distributions in place of uniform
priors on many of the model's parameters.
Using simulations, I show that the new implementation has improved robustness,
accuracy, and power compared to the original model.
The results confirm that the improved performance of the new model
is due to a combination of
(1) more flexible priors on divergence-time and demographic parameters that
avoid placing high prior density in improbable regions of parameter space, and
(2) a diffuse Dirichlet-process prior that does not strongly disfavor
divergence models with intermediate numbers of divergence events.
% Also, our results confirm that both the uniform priors on demographic and
% divergence-time parameters and the U-shaped prior on divergence models
% contribute to the tendency of \msb.
After reanalyzing sequence data from 22 pairs of taxa from the Philippines
\cite{Oaks2012} under the new model, I find a large amount of posterior
uncertainty in the number of divergence events shared among the taxa; a result
in contrast with the original \msb model and congruent with
intuition given the richness of the model and the relatively small amount of
information in the data.

@@

\subsection{The model}
In this section, I describe the model, which is a modification of the model
implemented in \msb \cite{Huang2011,Oaks2012}.
The code implementing the new model is freely available in the open-source
software package \dppmsbayes
(\href{https://github.com/joaks1/dpp-msbayes}{\url{https://github.com/joaks1/dpp-msbayes}}).
To perform the analyses described below, I used the freely avaliable,
open-source software package \pymsbayes
(\href{https://github.com/joaks1/PyMsBayes}{\url{https://github.com/joaks1/PyMsBayes}}),
which provides a multi-processing interface to \msb and
\dppmsbayes.
I performed the work described below following the principles of
Open Notebook Science.
Using version-control software, I make progress in all aspects of the work
freely and publicly available in real-time at
\href{https://github.com/joaks1/msbayes-experiments}{\url{https://github.com/joaks1/msbayes-experiments}}.
All information necessary to reproduce my results is provided there.
I follow much of the notation of Oaks et al.\ \cite{Oaks2012}, but modify it to
aid in the description of the new model.
A summary of my notation can be found in Table \ref{tabNotation}.

@@

\begin{linenomath}
To avoid calculating the likelihood terms of Equation \ref{eq:fullModelCompact},
I distill each sequence alignment \alignment{}{} into a vector of insufficient
summary statistics \alignmentSS{}{}, thus replacing the full dataset
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
with vectors of summary statistics for each alignment
$\ssVectorObs = (\alignmentSSObs{1}{1}, \ldots,
    \alignmentSSObs{\npairs{}}{\nloci{\npairs{}}})$.
\change{
Optionally, for each population pair, the means of the summary statistics can
be calculated across the \nloci{} loci, and the vector can be further reduced
to $\ssVectorObs = (\alignmentSSObs{1}{}, \ldots,
\alignmentSSObs{\npairs{}}{})$.
With \ssVectorObs in hand, we can estimate the approximate joint
posterior distribution}
{\small
\begin{equation}
    p(\geneTreeVector, \divTimeMapVector, \demographicParamVector, 
    \locusMutationRateScalarVector, \locusRateHetShapeParameter \given
    \ssSpace, \hkyModelVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) =
    \frac{p(\ssSpace \given \geneTreeVector, \hkyModelVector)
        p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
        \locusMutationRateScalarVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)
        p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter)
        p(\locusRateHetShapeParameter)
        p(\divTimeMapVector)
        p(\demographicParamVector)}{
        p(\ssSpace \given \hkyModelVector, \ploidyScalarVector,
        \mutationRateScalarConstantVector)},
    \label{eq:approxModelCompact}
\end{equation}
}
where \ssSpace is the multidimensional Euclidean space around the vector of
summary statistics, the radius of which is the tolerance \tol.
The sources of approximation are the insufficiency of the statistics and the
\tol being greater than zero.
I describe the full model in detail before delving into the numerical
method of estimating the approximate model.
\end{linenomath}

% \subsubsection{Likelihood terms of Equation \ref{eq:fullModelCompact}}
\change{
\subsubsection{Likelihood and gene-tree prior terms of Equation \ref{eq:fullModelCompact}}
}
\begin{linenomath}
% The likelihood terms of Equation \ref{eq:fullModelCompact} can
\change{
The likelihood and gene-tree prior terms of Equation \ref{eq:fullModelCompact}
}
can be expanded out as a product over population pairs and loci
\begin{equation}
    p(\alignmentVector \given \geneTreeVector, \hkyModelVector)
    p(\geneTreeVector \given \divTimeMapVector, \demographicParamVector,
    \locusMutationRateScalarVector, \ploidyScalarVector,
    \mutationRateScalarConstantVector) = 
    \prod_{i=1}^{\npairs{}}
    \prod_{j=1}^{\nloci{i}}
    p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})
    p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
    \locusMutationRateScalar{j}, \ploidyScalar{i}{j},
    \mutationRateScalarConstant{i}{j}).
    \label{eq:modelLikelihoodExpanded}
\end{equation}
The first term,
$p(\alignment{i}{j} \given \geneTree{i}{j}, \hkyModel{i}{j})$, is the
probability of the sequence alignment of locus $j$ for population pair $i$
given the gene tree and \hky \cite{HKY} substitution model parameters
\cite[i.e., the ``Felsenstein likelihood'']{Felsenstein1981}.
The model allows for an intra-locus recombination rate \recombinationRate,
which, for simplicity, is assumed to be zero in Equation
\ref{eq:fullModelCompact}.
If \recombinationRate is non-zero, this term requires an additional product
over the columns (sites) of each sequence alignment to allow sites to have
different genealogies.
The second term,
p(\geneTree{i}{j} \given \divTimeMap{i}, \demographicParams{i},
\locusMutationRateScalar{j}, \ploidyScalar{i}{j},
\mutationRateScalarConstant{i}{j}),
is the probability of the gene tree under a multi-population coalescent model
(i.e., species tree) where the ancestral population of pair $i$ diverges
and gives rise to the two sampled descendant populations.
Each \demographicParams{} contains the following demographic parameters: The
mutation-rate-scaled effective sizes ($\myTheta{} = 4N\mutationRate$) of the
ancestral, \ancestralTheta{}, and descendant populations, \descendantTheta{1}{}
and \descendantTheta{2}{};
the proportion of the first, \bottleScalar{1}{}, and second population,
\bottleScalar{2}{}, that persist during bottlenecks that begin immediately
after divergence in forward-time;
the proportion of time between present and divergence when the bottlenecks
end for both populations, \bottleTime{};
and the symmetric migration rate between the descendant populations,
\migrationRate{}.
Thus, the probability of the $\popSampleSize{i}{}-1$ coalescence times (node
heights) of gene tree \geneTree{i}{j} is given by a multi-population
Kingman-coalescent model \cite{Kingman1982} where the ancestral population of
size
$\ancestralTheta{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
diverges at time \divTimeMap{i} into two descendant populations of constant
size
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{1}{i}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}\bottleScalar{2}{i}$,
which, after time $\divTimeMap{i}\bottleTime{i}$, grow exponentially to their
present size 
$\descendantTheta{1}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$
and
$\descendantTheta{2}{i}\ploidyScalar{i}{j}\mutationRateScalarConstant{i}{j}\locusMutationRateScalar{j}$,
respectively.
Following divergence, the descendant populations of pair $i$ exchange migrants at a symmetric rate of
\migrationRate{i}.
\end{linenomath}

% \subsubsection{Prior terms of Equation \ref{eq:fullModelCompact}}
\change{
\subsubsection{Additional prior terms of Equation \ref{eq:fullModelCompact}}
}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate multipliers
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be non-zero, the
prior term $p(\recombinationRate)$ would be added to Equation
\ref{eq:fullModelCompact}, and the prior would be $\recombinationRate \sim
Gamma(\gammaShape{\recombinationRate}, \gammaScale{\recombinationRate})$,
\change{where \gammaShape{\recombinationRate} and
\gammaScale{\recombinationRate} are specified by the investigator.}
\end{linenomath}

\begin{linenomath}
The prior term for the demographic parameters, $p(\demographicParamVector)$,
expands out into its components and as a product over the \npairs{}
pairs of populations
\begin{equation}
    p(\demographicParamVector) =
    \prod_{i=1}^{\npairs{}}
    p(\ancestralTheta{i})
    p(\descendantTheta{1}{i})
    p(\descendantTheta{2}{i})
    p(\bottleScalar{1}{i})
    p(\bottleScalar{2}{i})
    p(\bottleTime{i})
    p(\migrationRate{i}).
    \label{eq:demographicPrior}
\end{equation}
\change{
The priors for the demographic parameters are
$\ancestralTheta{} \sim Gamma(\gammaShape{\ancestralTheta{}},
\gammaScale{\ancestralTheta{}})$,
$\descendantTheta{1}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\descendantTheta{2}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
$\bottleScalar{1}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleScalar{2}{} \sim Beta(\betaA{\bottleScalar{}{}},
\betaB{\bottleScalar{}{}})$,
$\bottleTime{} \sim U(0, 1)$,
and
$\migrationRate{} \sim Gamma(\gammaShape{\migrationRate{}},
\gammaScale{\migrationRate{}})$,
where the hyper-parameters of each prior distribution can be specified by
the investigator.
By default, \ancestralTheta{}, \descendantTheta{1}{}, and \descendantTheta{2}{}
share the same prior (i.e., $\gammaShape{\ancestralTheta{}} =
\gammaShape{\descendantTheta{}{}}$ and $\gammaScale{\ancestralTheta{}} =
\gammaScale{\descendantTheta{}{}}$), but a separate gamma-distributed prior can
be assigned to \ancestralTheta{}.
Also, the \bottleScalar{1}{}, \bottleScalar{2}{}, and \migrationRate{}
parameters are optional (i.e., the investigator can assume that there has been
no migration between populations of each pair and/or the population size of
each descendant population has been constant through time).}
\end{linenomath}

@@

\subsubsection{Time scale}
As described above, divergence times are in units of
$\myTheta{C}/\mutationRate$ generations, where \myTheta{C} is the expectation
of the prior on descendant-population size.
As described by Oaks et al.\ \cite{Oaks2012}, in \msb, \myTheta{C} is half of
the upper limit of the continuous uniform prior on the mean of the descendant
population sizes.
This is only equal to the expectation of the prior if the lower limit of the
prior is zero.

\subsection{ABC estimation of the posterior of the model}
\subsubsection{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
I use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
\change{
(1) drawing values of all the model's parameters, which I denote \hpvector{},
from their respective prior distributions;
% (2) rescaling the divergence times
% $\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
% from units proportional to time to units proportional to the expected number
% of mutations via Equation \ref{eq:divTimeScaled} to get 
% $\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
% \divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(2) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(3) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
and
(4) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments.
Optionally, an additional step can be performed to reduce the summary
statistics to the means across loci for each population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$.
Either way, \ssVector{} contains the same summary statistics as those estimated
from the observed data \ssVectorObs.
}
After repeating this procedure \numPriorSamples times, we have a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.
